Lost in the woods: shifting habitats can lead phylogeography astray

### 2025-03-30 manuscript/ResponseToReviews.tex

\response{
Thank you for the opportunity to revise our manuscript and for the thoughtful comments from the reviewers.
I particularly appreciate the detailed constructive feedback from Simon Dellicour.
I hope I have addressed all points in a satisfactory manner and that this manuscript will be a valuable contribution to the discussions of the value and pitfalls of phylogeographic inference.

Below are point-by-point responses to each reviewer.
}

\section*{Reviewer 1 comments}

{\bf Comments for the Authors.}
In this study, the author rigorously critiques continuous phylogeographic models and shows how their posterior inferences are unreliable due to shifts/changes in the habitats or environmental conditions, particularly in poorly sampled areas. I enjoyed reading this well-written paper and analyses, and I might agree with many points raised by the author. However, I have also issues with his illustration summarized in the following points:

@@

\response{Thank you very much for this assessment and for highlighting where the manuscript falls short. I address these points individually below.}

{\bf Major concerns:}

1. He is basing his conclusions on simulated data, not actual examples of real sequences for specific species. In fact, I think the author needs to illustrate his points using sequence data from different species.

\response{The purpose of this manuscript is to show that the assumptions of continuous phylogeographic methods can be problematic in several ways. To show that, I need scenarios where the ground truth is known and this is not the case for observational data.
Furthermore, we are not proposing an alternative method to analyze data -- so it is unclear to me what exactly the reviewer is asking me to do.
I discuss the expansion of West Nile virus in the US as an example where the assumptions of continuous phylogeographic models are likely violated and show how a different set of assumptions leads to dramatically different conclusions.
}

@@

2. Also, it is very important to note that most of the published results from phylogeographic models do exhibit acceptable epidemiological plausibility to biology and the distribution of species in space and time.

\response{I discuss this towards the end of the manuscript. Phylogeographic methods tend to produce a parsimonious reconstruction of past migrations which is often plausible (and will often also cover the actual history). However, these inferences depend on certain assumptions about the past and extrapolations into the past are dangerous.

I have extended this section in the discussion to make clear when plausible inferences are expected, and when inferences should interpreted with care.
}

@@

3. Also, the journal is specialized in virus evolution, so it is important to shift the scope of the paper to the implications of such quantitative findings to the biology and genomics of viruses or similar rapidly evolving pathogens.

\response{The methods that discussed in this paper are widely used in the analysis of viral sequences and often were developed in the context of viral evolution. While the implications of some of the issues might apply more broadly, I feel that phylogeographic inference is sufficiently central to the analysis of viral genomic data and the readership of Virus Evolution, that the current focus is justified.
I discuss examples of Rabius virus and West Nile virus. }

\section*{Reviewer 2 comments}
- general comment: given the popularity of phylogeographic approaches, especially in the field of molecular epidemiology to infer the dispersal history of viral lineages, the limitations addressed and explored in this study are of importance for the community of users of those approaches. In addition to a series of more minor comments and suggestions listed below, I would have a comment about a potential ambiguity on the possibility to apply an alternative approach based on a structured coalescent model that would allow to circumvent the coupling issue illustrated in the present study focusing on the continuous phylogeographic approach (see the next two comments). The focus of this study is indeed on the continuous diffusion model, and I think that this is important to underline that the author did not demonstrate that in practice the use of such a alternative approach would allow avoiding this issue (even if I'm not saying that he should, see my related comment just below). While a criticism of these limitations inherent to the current implementation of the continuous diffusion model implemented in BEAST are totally relevant, the other phylogeographic approaches/models are also associated with a number of assumptions and limitations (for instance, any outcome of phylogeographic inference will be impacted by sampling bias).

@@

- Abstract: in it, the author never mentioned that his investigations solely focused on the continuous phylogeographic inference. Imo, without this precision (that is explicitly stated in the first sentence of the second paragraph of the Introduction section), it could sound that most popular phylogeographic models/approaches were assess in regard to those limitations.

\response{I agree that it is important to make this point clear. The first word of the abstract already read `Continuous`, but I have added additional qualifiers to clarify that the focus is on continuous phylogeographic inference.}

- p1, Introduction section: some developments associated with or related to the continuous diffusion model are not mentioned or addressed in that section. For instance: Fitzjohn et al. (2009, doi: 10.1093/sysbio/syp067), Guindon et al. (2016, doi: 10.1016/j.tpb.2016.05.002), Gill et al. (2017, doi: 10.1093/sysbio/syw093).

\response{Thank you for pointing out these references. However, I could not see the relevance of Fitzjohn et al 10.1093/sysbio/syp067 to the present study. Fitzjohn et al study estimation of speciation rates in incompletely resolved phylogenies. I have added the reference Guidon et al when discussing challenges of defining consistent spatial models for inference. Gill et al seem mostly relevant in the context of directional random walks models. I have added a reference to this work when discussing range expansions.}

@@

- p2, "While the estimator Dw directly estimates a parameter of the model and its behavior has been studied in simulations (Pybus et al., 2012)": Pybus et al. didn't use any simulation in their study. Except if the author aimed to refer to a different study, in my knowledge, an analysis of the diffusion coefficient(s) behaviour based on simulations was also lacking (until recent related works mentioned above and the present study).

\response{Thank you for pointing out this error. I had confused myself. Pybus et al discuss how this estimator uses fundamental properties of diffusion -- with a reference to Einstein!. I reworded this sentence.}

- p3, "While the estimator Dw as defined in Eq. 2 is well behaved when the model is exactly diffusive, long range dispersal can affect Dw dramatically. If the probability of long jumps with distance $x < d < x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the Dw diverges": as it could be of interest for future works involving the estimation of $D_w$, I was wondering if the author could further develop the later statement to explain why this is actually expected. It is maybe obvious, but at least not for me, so, if relevant, a more detailed development could be inserted (?). Also, and my apologies if I missed something here, but how do you define a "long jump"?

\response{
The estimator $D_w$ is an empirical average over the squared displacement along different branches, scaled by the inverse elapsed time.
However, the Cauchy distribution is often used to model long range dispersal, but it does not have a finite mean nor a variance. The central limit theorem does not apply for sums of many such variables and estimators based on such sums will be dominated by the largest jumps. Estimates of $D_w$ will therefore be very noisy and often dominated by largest jumps possible in the systems.
In the concrete case of a Cauchy distribution, the largest jump will typically be twice the distance of the second largest jump, 3 times as large as the third largest jump, etc. The sum in the estimator $D_w$ is them dominated by the largest jump (and hence hugely variable).
I have added a footnote elaborating this issue.

I think broad distribution are fine as fairly uninformative priors for inference, but the estimate $D_w$ of a forward process with long jumps is not well defined.

}

- Figure 3: I had a hard time understanding it. I would for instance suggest to be explicit with what you mean by "rel. heterogeneity" on the y-axis of panel A (which is "the standard deviation of population density in two dimensional bins of size L/5"), or at least to re-state it in the figure's legend. Also, I would re-define, within the figure's legend, to what "r" refers to (and its unit?). Finally, in panel C, this is difficult to distinguish the curves and their associated symbols (symbols that could also be re-defined in a colour legend included within the box of the graph).

@@

- p3, "For $D < L2/2Tc$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals": maybe mark it with a vertical mark (/dashed line) in Figure 3?

\response{Thank you for these suggestions. I have renamed the y-axis to "density fluctuations" and explained how they are defined in more detail in the text. Their definition is also given in the caption. I have also specified in the caption that $r$ is the interaction radius of individuals (and thus is a length) and added a vertical line indicating $D=L^2/2T_c$.  }

@@

- Figure 4: a colour scale is missing in panel A, which I didn't find intuitive. Also, in the legend: "moved [to?] the left"?

\response{Thank you for pointing out the missing word. I didn't think that adding a color scale would help much, but I have added sentence to the caption detailing how yellow and purple correspond to the most extreme habitats.}

- p5, "This is expected as there is no information in the current sample to suggest otherwise, but it highlights the potential of misleading inference, in particular when extrapolating far into the past": (i) do you get the same trend while considering a RRW model? (ii) Is the process only sample at the end? (What happens if we have dated samples?) (iii) Without a complex time varying environment, a RRW model with drift (Gill et al. 2017) might potentially achieve the same kind of result where the root does not lie in between the samples.

\response{This statement refers explicitly to 'extrapolation' into the past and the inference in this case was made form data gathered at one time point. If there were samples from the time of the $T_{mrca}$ to the present, the task would be interpolation rather than extrapolation and I would expect the biases to disappear. Integrating over a distribution of diffusion constants will likely broaden the confidence intervals and thus make the estimates more consistent with reality, but I don't expect the bias to go away.}

@@

1.  ``This model assumes that the dispersal of one lineage is independent of other lineages, that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time. Furthermore, it assumes that the tree topology and branching rates are independent of spatial location.'' I agree with these points, but it would be helpful to explain why these assumptions are implied by the seemingly innocuous Eq. 1. Users of diffusion-based phylogeographic inference methods may not be aware that these assumptions are inherent.

\response{Thank you for this suggestion. I have expanded this section.}

2. The section on density regulation seems to assume global or habitat-wide density regulation. However, it is plausible that many species are density-regulated over a characteristic length scale, such as their dispersal distance during their lifetime (or from birth to reproductive age). In that case, wouldn’t the parameter  L  (habitat size) be replaced with the scale over which density is regulated? If so, this should be mentioned. A static, homogeneous, and locally density-regulated population should then approximately follow Eq. 1.

\response{In the individual based simulation, fitness of individuals is determined by a local density estimator which is given by a sum of Gaussians with characteristic radius $r$ for each individual in the population (see Eq(4)). However, even when local density fluctuates much less, the population still deviates from the assumption of the typical phylodynamic/phylogeographic model since the rate of coalescence between any two lineages strongly depends on their spatial separation.}

3. ``which cab lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low'': Replace ``cab'' with ``can''. Additionally, in this context, beyond Wilkins and Wakeley, it may be worthwhile to cite the Appendix B of Hallatschek, O., \& Nelson, D. R. (2008). Gene surfing in expanding populations. Theoretical population biology, 73(1), 158-170.

@@

\response{Done.}

4. It might be helpful to provide guidance on whether users can make quantitative estimates or perform tests to assess the reliability of their phylogeographic inferences. For example, how dense does a time series need to be to sufficiently constrain the spatiotemporal embedding? Alternatively, this manuscript could call for studies that address this issue. Following this work, the onus is on the phylogeographic community to develop best practices for these methods.

\response{Thank you for this suggestion. My gut feeling in this context is "interpolation is mostly fine, extrapolation problematic", while parameter estimates are typically hard to interpret. I have elaborated on this a bit in the final paragraph of the discussion.}

### 2025-03-30 manuscript/phylogeo.tex
work on response

In either case, estimates of $v_w$ or $v_b$ are not meaningful.
The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity.
There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted \citet{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the behavior using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.

While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, long range dispersal can affect $D_w$ dramatically.
If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
In this case, the values of the above estimators for a dataset with long range dispersal are dominated by the longest displacements and a single number like $D_w$ is no longer a useful summary of the typical dispersal process.
Estimators like $D_w$ might also become problematic under the assumptions of relaxed random walk models with long-tailed prior distributions for $D$.

\section*{Effects of density regulation on population structure and dispersal}
Continuous phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}.
If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of a panmictic population ($T_c=N$ in the case of a neutral model with a population of size $N$).
Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of size $L/5$ relative to the expectation in the case of a spatially well-mixed population when each individual picks its location at random.
For $D< L^2/2T_c$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals.
This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
The comparison of $D$ to the ratio $L^2/T_c$ will continue to be important below, and I therefore use diffusion constants in units of $L^2/T_c$.

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/stable_density}
    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat. The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii. For $r\ll L$, density fluctuations are strongly suppressed.
    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
    }
\end{figure}

@@

Phylogeographic methods are often studied using the spread of West Nile virus in North America and Rabies virus.
West-Nile virus was first detected on the East Coast of the US in 1999 and reached the West Coast five years later in 2004 \citep{pybus_unifying_2012}, translating to a wave front velocity of about 1000 km/year.
\citet{pybus_unifying_2012} estimated a diffusion coefficient between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
These extremely high estimates of diffusive dispersal are implausible.
Alternatively, one can interpret the spread from east to west as a range expansion.
To this end, we need an estimate of the growth rate $\alpha$ of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain estimates of $D$ of $D\approx \frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
This lower estimate of $D$ implies a daily exploration radius of about 5 km.
This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020}: Once the habitat was fully explored, directed range expansion with $\vfkpp$ ceases.
Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks \citep{hallatschek_acceleration_2014}.

@@

Why does phylogeography then often generate sensible results?
On short time scales, a diffusive model will mostly reduce to a parsimonious reconstruction of ancestral locations which is robust to details of the model.
Sampling at multiple time points further constrains ancestral locations sufficiently strongly that there is limited uncertainty, at least within the period for which samples are available.
But inferences for deeper nodes in the phylogeny before the earliest samples can be problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial information, they can also be highly unreliable and misleading.
In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless fail to faithfully infer history or essential parameters.

>>>>

In either case, estimates of $v_w$ or $v_b$ are not meaningful.
The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity.
There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted \citet{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the sample size dependence using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.

While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, long range dispersal can affect $D_w$ dramatically.
If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
When jumps are drawn from a Cauchy distribution, as for example used in \citet{dellicour_how_2024}, the estimator $D_w$ will be dominated by the longest jumps (or the size of the system)\footnote{The largest jump in a sample of $n$ jumps is typically on the order of $2\gamma n/\pi$, where $\gamma$ is the scale of the Cauchy distribution. The second largest jump is typically half, the third largest a third, etc.~of the largest value. The estimator $D_w$ sums their squares, which in many cases will be dominated by a few very largest values. }.
In this case, $D_w$ is no longer a useful summary of the typical dispersal process as it is determined by the most extreme events that are will fluctuate widely between different realizations of the process or even different samples from the same realization.

\section*{Effects of density regulation on population structure and dispersal}
Continuous phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}:
The clades defined by long branches in the tree correspond to tight clusters of individuals in space.
A more realistic population would spread out more evenly across the habitat.

If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of a panmictic population ($T_c=N$ in the case of a neutral model with a population of size $N$).
I quantified this heterogeneity using by binning individuals into two-dimensional bins with linear dimension $L/5$ (25 bins in total) and calculated the standard deviation of the number of individuals in these bins, normalized by the expectation variation in a well-mixed population when each individual picks its location at random.
The black line in Fig.~\ref{fig:density_reg}A shows this heterogeneity measured as this standard deviation for a purely diffusive model as a function of the diffusion constant.
For $D < L^2/2T_c$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals and high heterogeneity.
This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
The comparison of $D$ to the ratio $L^2/T_c$ will continue to be important below, and I therefore use diffusion constants in units of $L^2/T_c$.

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/stable_density}
    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat. The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii $r$. For $r\ll L$, density fluctuations are strongly suppressed.
    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
    The dashed vertical line indicates $D=L^2/2T_c$.}
\end{figure}

@@

Phylogeographic methods are often studied using the spread of West Nile virus in North America and Rabies virus.
West-Nile virus was first detected on the East Coast of the US in 1999 and reached the West Coast five years later in 2004 \citep{pybus_unifying_2012}, translating to a wave front velocity of about 1000 km/year.
\citet{pybus_unifying_2012} estimated a diffusion coefficient between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
\citet{dellicour_how_2024} estimated a diffusion constant of 310$km^2$/day for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
These extremely high estimates of diffusive dispersal are implausible.
Alternatively, one can interpret the spread from east to west as a range expansion.
To this end, we need an estimate of the growth rate $\alpha$ of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain estimates of $D$ of $D\approx \frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
This lower estimate of $D$ implies a daily exploration radius of about 5 km.
This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020, dellicour_how_2024}: Once the habitat was fully explored, directed range expansion with $\vfkpp$ ceases.
Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks \citep{hallatschek_acceleration_2014}.
West Nile virus might for example by carried by migratory birds, but their directed migration should not be described by a diffusion constant.

@@

Why does phylogeography then often generate sensible results?
On short time scales, a diffusive model will mostly reduce to a parsimonious reconstruction of ancestral locations which is robust to details of the model.
Sampling at multiple time points further constrains ancestral locations sufficiently strongly that there is limited uncertainty, at least within the period for which samples are available.
In these cases, phylogeographic reconstruction effectively interpolates been sampled locations along the relationship structure encoded in the tree and the results tend to be both plausible and well constrained by the data.
But inferences for deeper nodes in the phylogeny long before the earliest samples can be problematic as the chance that the environment has changed and shifted, or that the true ancestral locations are in poorly sampled regions, increases.
And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial information, they can also be highly unreliable and misleading.
In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
This concern also applies to scenarios where the spatial component of the likelihood strongly influences the posterior of phylogenetic or phylodynamic parameters. In such cases, chances are that model misspecification such as missing interaction between spatial location and population dynamics will lead to biased inferences. Integrating over broad distributions of diffusion constants will make the model more permissive to absorb such misspecifications, but does not address the underlying problem.

In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless fail to faithfully infer history or essential parameters.
